---
title: "02_basics"
author: "Emily Kim"
date: "2025-01-20"
output: 
  html_document:   
    theme: united  
    toc: true  
    toc_float: true  
    keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
     
https://daattali.gitbooks.io/stat545-ubc-github-io/content/cm002_r-rstudio-intro.html  
https://daattali.gitbooks.io/stat545-ubc-github-io/content/block002_hello-r-workspace-wd-project.html  

## Update R and R Studio: 
https://happygitwithr.com/install-r-rstudio  https://happygitwithr.com/rmd-test-drive   
Update RStudio every 1-2 months:  check version `RStudio > About RStudio`      
Update R 1-2x / year, last major change in April 2020 with 4.0.0:  `R.version.string`
Stay < 1 minor version behind, else pre-built binary add-on packages from CRAN will not install   
rmarkdown:  check version `packageVersion("rmarkdown")` 

## RStudio
Default panes:  

* Console (left)  
* Environment/History (upper right)  
* Files/Plots/Packages/Help (lower right) 

Keyboard shortcuts: `option`+`shift`+`K`

Quit RStudio: ⌘Q or `q()` in Console  
Save workspace: `Save workspace image to ~/.Rdata`  

## R Objects
Create objects with the operator `<-`  

* `option` + `-` in Console    
* "objectNames gets value" - give the object an assignment    
```
objectName <- value
```
Object names:   

* Can NOT start with a digit. 
* Can NOT contain a comma or space.  

Demarcate words:  

* snake_case
* use.periods
* CamelCase

## R functions 
Specify a function: `argument = value` or argument by position  
`functionName(arg1 = val1, arg2 = val2, and so on)`

Functions: Type part of name and hit TAB. Floating help shows arguments. F1 shows full documentation.  
`seq()` makes sequences of numbers  

1) Type `se` and hit `Tab` to see possible completions 
2) Specify `seq` by typing more or using up/down arrows  
3) Type opening parenthesis:  auto adding of closing parenthesis, cursor placed in middle  
4) Type `1,10` and hit return to exit the expression.   
5) R assumes `from = 1` and `to = 10` and `step = 1`. 
```
seq(1,10)
```
Surround the assignment with parentheses to "print to screen"  
```
(y <- seq(1,10)))
```
Some functions don't require arguments 
```{r}
date()
```

## Workspace

Objects appear in the workspace Environment

* List objects: `objects()` or `ls()` 
* Remove object y: `rm(y)`
* Remove everything: `rm(list=ls())`   
* or click the broom in Environment pane  

## Source code  
Save your R scripts: reproduce your analysis with input data and R code.     

Place source code for every object in a particular directory for later editing and retrieval. 
Write code for humans, write data for computers  
Files names: number_clear-description.filetype  
Dates: yyyy-mm-dd_clear-description.filetype  

## Working directory  
Organize your projects into directories. Set the working directory to the project directory.  

Where R looks for files to load. Where R writes files to disk.  
Check working directory or set working directory  
```r
getwd()
setwd("~/ekim/R/repo/R/stat545")
```

## RStudio Project
All project files - input data, R scripts, analytic results, figures  
Open project to set working directory  

Enter some commands in the Console:
```
a <- 2
b <- -3
sig_sq <- 0.5
x <- runif(40)
y <- a + b * x + rnorm(40, sd = sqrt(sig_sq))
(avg_x <- mean(x))
```
Plot result
```
write(avg_x, "avg_x.txt")
plot(x, y)
abline(a, b, col = "purple")
```
Save output as pdf
```
pdf("my_plot.pdf") # open device
plot(x,y)
abline(a,b, col = "purple")
dev.off() 
```
This is a good start, so let's preserve the logic and code.  
`History` > `To Source` to paste the code into `Source` as a new R script     
Save the code as R script `.R` or `.r`  or Rmd `.Rmd`  
`#` to comment `command + shift + C`  
Quit and inspect project folder. View the PDF.  

Restart RStudio  
Edit code  

* Set the sample size `n` at the top, replace hard-wired 40's with `n`  
* Alter the sample size `n`, the slope of line `b`, the color of the line  

Re-run code

* Walk through line-by-line `command + enter` 
* Run several lines `Run` > `Run Selected Line(s)`  
* Source the entire `.Rmd`
  * `shift`+`command`+`S` or  `Run` > `Run All`
* Source entire `.R`  
  * `source('toy-script.r')` in Console  
  * mouse > Source  
  
Visit your figure in an external viewer to verify that the PDF is changing as expected  
Find the script that created the PDF  

```
# Add {r} to this code to run  
n <- 40 # set sample size
a <- 2
b <- -3
sig_sq <- 0.5
x <- runif(n)
y <- a + b * x + rnorm(n, sd = sqrt(sig_sq))
(avg_x <- mean(x))

pdf("my_plot.pdf",         # File name
    width = 8, height = 7, # Width and height in inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
plot(x,y)
abline(a,b, col = "purple")
dev.off()
```

## Workflow

* Create an RStudio project for an analytical project  
* Keep inputs there (for importing)     
* Keep scripts there; edit them, run them in bits or as a whole from there
  * Load data or save a figure using code, not the mouse  
* Keep outputs there (like the PDF written above)  

Clean workspace with broom or `rm(list=ls())`  
Restart R to re-run analysis to ensure that code is complete/correct     





 